A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2022-09-18 submission/manuscript.rmd
latest updates for submission

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
author: 
  - Alexandros Rekkas, MSc${^1}$ 
  - David van Klaveren, PhD$^{2,3}$, 
  - Patrick B. Ryan, PhD$^4$
  - Ewout W. Steyerberg, PhD$^{3,5}$
  - David M. Kent, MD$^2$
  - Peter R. Rijnbeek, PhD${^1}$
output:
    bookdown::pdf_document2:
      number_sections: false
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage[labelfont=bf]{caption}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: the-lancet.csl
---

@@

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 8069six. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

@@

```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
include_graphics(here("figures/rsee_process.png"))
```

@@

For our demonstration we used data from three US claims databases, namely IBM
MarketScan Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid
(MDCD), and IBM MarketScan Medicare Supplemental Beneficiaries (MDCR). Our
analyses included a total of
`r patientNumbers %>% filter(database == "ccae") %>% pull(treatment)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(treatment)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(treatment)`
patients initiating treatment with ACE inhibitors and 
`r patientNumbers %>% filter(database == "ccae") %>% pull(comparator)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(comparator)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(comparator)`
patients initiating treatment with beta blockers in CCAE, MDCD and MDCR
respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients
were included in all strata of predicted acute MI risk (Supplement, page 26).

@@

The models had moderate discriminative performance (internally validated) with
no major issues of overfitting to any cohort except for the case of CCAE, where
the derived prediction model performed better in the comparator cohort (Table
\@ref(tab:predictionPerformance)). We also observed lower performance of the
prediction model developed in MDCR compared to the other 2 databases
(Supplement, page 29).

@@

Finally, the distribution of the estimated relative risks with regard to 30
negative control outcomes indicated unresolved confounding within the lowest
acute MI risk quarter of CCAE (Figrue \ref{fig:negativeControls}). Hazard ratios
significantly different than 1 (true effect size) were concentrated in the lower
right part of Figure \ref{fig:negativeControls}: panel Q1. This suggests
significant negative effects of ACE inhibitors compared to beta blockers on
causally unrelated outcomes, pointing at unresolved differences between the two
treatment arms. This was not the case in the other risk quarters of CCAE, or in
any risk quarter of MDCD and MDCR (Supplement, pages 34-36).

@@

# CODE AVAILABILITY
Our analysis code is publicly available under the MIT license at
https://github.com/mi-erasmusmc/AceBeta9Outcomes.

@@

>>>>

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
author: 
  - Alexandros Rekkas, MSc${^1}$ 
  - David van Klaveren, PhD$^{2,3}$, 
  - Patrick B. Ryan, PhD$^4$
  - Ewout W. Steyerberg, PhD$^{3,5}$
  - David M. Kent, MD$^2$
  - Peter R. Rijnbeek, PhD${^1}$
output:
    bookdown::pdf_document2:
      number_sections: false
      keep_tex: true
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage[labelfont=bf]{caption}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: the-lancet.csl
---

@@

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

@@

```{r graphicalAbstract, cache=TRUE, echo=FALSE, fig.cap="\\textbf{Framework execution diagram}. Illustration of how the framework is applied on two observational databases, preferably mapped to OMOP-CDM.", fig.show="hold", out.width = '100%'}
include_graphics(here("figures/rsee_process.png"))
```

@@

For our demonstration we used data from three US claims databases, namely IBM
MarketScan Commercial Claims and Encounters (CCAE), IBM MarketScan Medicaid
(MDCD), and IBM MarketScan Medicare Supplemental Beneficiaries (MDCR). Our
analyses included a total of
`r patientNumbers %>% filter(database == "ccae") %>% pull(treatment)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(treatment)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(treatment)`
patients initiating treatment with ACE inhibitors and 
`r patientNumbers %>% filter(database == "ccae") %>% pull(comparator)`,
`r patientNumbers %>% filter(database == "mdcd") %>% pull(comparator)`, and
`r patientNumbers %>% filter(database == "mdcr") %>% pull(comparator)`
patients initiating treatment with beta blockers in CCAE, MDCD and MDCR
respectively (Table \@ref(tab:incidenceOverall)). Adequate numbers of patients
were included in all strata of predicted acute MI risk (Supplementary Tables
S1-S3).

@@

The models had moderate discriminative performance (internally validated) with
no major issues of overfitting to any cohort except for the case of CCAE, where
the derived prediction model performed better in the comparator cohort (Table
\@ref(tab:predictionPerformance)). We also observed lower performance of the
prediction model developed in MDCR compared to the other 2 databases.

@@

Finally, the distribution of the estimated relative risks with regard to 30
negative control outcomes indicated unresolved confounding within the lowest
acute MI risk quarter of CCAE (Figrue \ref{fig:negativeControls}). Hazard ratios
significantly different than 1 (true effect size) were concentrated in the lower
right part of Figure \ref{fig:negativeControls}: panel Q1. This suggests
significant negative effects of ACE inhibitors compared to beta blockers on
causally unrelated outcomes, pointing at unresolved differences between the two
treatment arms. This was not the case in the other risk quarters of CCAE, or in
any risk quarter of MDCD and MDCR (Supplementary Figures S1-S2).

@@

# CODE AVAILABILITY
The source code for the R-package that implements our framework can be found at
https://github.com/OHDSI/RiskStratifiedEstimation. The code for implementing the
proof of concept study presented here is publicly available at
https://github.com/mi-erasmusmc/AceBeta9Outcomes.

@@

# ACKNOWLEDGEMENTS
AR and PRR have received funding from the Innovative Medicines Initiative 2
Joint Undertaking (JU) under grant agreement No 80696. The JU receives support
from the European Union’s Horizon 2020 research and innovation programme and
EFPIA.
